### 2020-03-08 thesis/paper.tex
wording improvement

Clustering is the task of finding groups of points in data such that points within a group have some kind of relationship, while points in different groups do not relate. Defining how such a relationship can be expressed and how these groupings can be found are non-trivial tasks, for which reason a lot of research has been put towards this topic. Different ways of thinking about this problem, the expected results and the performance aspects of the algorithms resulted in a large number of different methods.

 To summarize which methods are available many surveys were written over time \cite{7154919,1427769,7414675,surveyclustering}. Another aspect of clustering is which kinds of data can be clustered and even for more specific tasks like text clustering many methods are available \cite{5982288}. Even though these are relevant methods, they will not be mentioned further on as this paper only works on processing numeric data. Also, work has been put into optimizing clustering algorithms for different types of architectures and machines, e.g. for distributed systems \cite{6322592} or optimized for energy consumption and load balancing \cite{7586361}. Such aspects will also not be further considered as this paper and the created tool do not aim to beat existing methods in runtime or energy efficiency.

To give an overview of existing clustering algorithms Xu and Tian \cite{surveyclustering} generalize the idea of algorithms to nine categories. These categories are defined by which fundamental ideas the algorithms are based on, which are the following:
\begin{itemize}
  \item Partition
  \item Hierarchy
  \item Fuzzy Theory
  \item Distribution
  \item Density
  \item Graph Theory
  \item Grid
  \item Fractal Theory
  \item Model
\end{itemize}

Partition based methods aim to find central points in the data to represent clusters. Data points are then assigned to the central point closest to them, whereby each central point forms a cluster. The most famous method in this group is k-Means which has been implemented in different ways, an example for a newer variant of it was proposed by Kanungo et al. \cite{1017616}.

Hierarchy based methods at first consider each point as representing a cluster and merge groups of points to form clusters according to some logic. The opposite can also be done, first assuming all points belong to one cluster and the spiting them into multiple clusters. When starting with each point in one cluster, typical linking strategies are single link, joining clusters according to the distance between the closest points, average link, joining according to the average distance of all points from one cluster to the other, or maximum link, whereby the maximum distance between points of the clusters is considered.

Fuzzy Theory based methods do unlike other methods not produce discrete values for the cluster ownership label but a value in the interval $[0,1]$ representing how strongly a point belongs to a cluster. The usual method to compute such an ownership value is to optimize toward an objective function and seeing these values as the likelihood of a point belonging to a given cluster.

Distribution based methods aim to find clusters with the idea that any cluster should have its own mathematical distribution. In other words, each distribution found in the data should have a cluster with its generated points assigned to it. In most cases Gaussian distributions are assumed, reducing the complexity of the search.

@@

Graph Theory based methods work by regarding each point in the data as a vertex of a graph with their distances representing the edges. The algorithms then aim to partition the graph into $k$ sub-graphs each representing a cluster, whereby the edges connecting the sub-graphs should have a large sum of the distances or small sum of similarities.

Grid based methods look at the data space as a multidimensional grid. The resulting rectangular spaces are then used in combination with hierarchical and density based methods to both improve the result and the computational performance of the algorithms. The size of the cells of the grid allows a trade-off between those two goals

Fractal Theory based methods aim to optimize the intrinsic quality of the fractal dimension \cite{surveyclustering}. In this context, fractal represents the geometry divisible into multiple parts with common characters with the whole \cite{surveyclustering}.

@@

%more on current development?
\section{Visual Frameworks}
Visual Frameworks are used to both allow the user to analyze data and empower him to produce better results. Often, when tackling a data analysis problem, the issue is not that there are no good methods available to solve it but just that the choice of which method produces a satisfying result is difficult. As looking at quality values and data tables does often not reveal the necessary information for educated choices, visualizations and frameworks are needed. Note that for this paper the words framework, tool and application will be used synonymously as I consider them only differing in the way how the user uses the created work, whether it is to be used from code, through APIs or just as a program with a graphical user interface.

@@

For this reason, different frameworks have been proposed in literature, a prominent one of those being Clustervision \cite{Kwon2018ClustervisionVS}. The Clustervision framework allows the user to visually explore the solutions created by multiple clustering algorithm runs, whereby each run may use a different algorithm and different settings for computation. Besides just showing the result and information on different properties of the resulting clusters, this framework ranks the computed results according to quality measures. As there exist a large number of different quality measures, each being biased towards/preferring specific cluster structures, the framework uses the average of five different quality measures as an indicator of the quality. A problem of this approach may be that when individual measures are biased and are combined into a more general one, one must be careful to not create new biases by the choice of how to combine them. The choices made by the authors of Clustervision \cite{Kwon2018ClustervisionVS} may for example show favorable results for the presented use-cases but may be poor when facing a data-set that does not fit the assumptions of the chosen metrics. Aside from the problems arising from quality measures, Clustervision only considers clustering solutions that are at least $5\%$ different from other top results, possibly missing better solutions because they are similar to a good result. Also, by choosing a single result through this ranking process the final choice can in the best case be as good as the best produced clustering, even though much more information is available when looking at all solutions. This left out knowledge can possibly be used to produce a combined result, that may be better than any individual one resulting from simple clustering algorithms. These points are also the reason why the framework described later in this thesis was created.

VISTA \cite{VISTA} is another visual clustering framework which was proposed by Chen and Liu and consists of three parts. Firstly, VISTA allows for data filtering which prepares data for visualization. In this step data can be normalized, missing values can be handled and if the dimensionality is too high, reduction techniques can be used. Also, if the data is too large it is possible to reduce it by extracting relevant sub-sets. Next, selecting and filtering of labels is possible. After clustering, it is possible to partially extract cluster labels for the points using the framework's data filtering tools, allowing to for example only visualize some clusters. Lastly, the tool allows for post-processing whereby a new cluster presentation is used which they name ClusterMap. ClusterMap projects the data on a two-dimensional grid showing the cluster labels in each cell and allowing for quick relabeling. Chen and Liu further iterate on this idea by introducing iVIBRATE \cite{10.1145/1148020.1148024} whereby they further analyze the accuracy of their tool and the theoretic background of their method.

Many clustering frameworks also implement visual elements such that clustering results or data can be explored visually, although this often only includes trivial plots. For example, the Java framework ELKI provides such simple functionality \cite{10.14778/2824032.2824115}. Generally, in most programming languages simple scatter plots and charts are available, which can be used whenever no further in-depth analysis is needed.

\subsubsection{Data-set specific Frameworks}
On top of the general frameworks for visualizing clusterings, specific ones exist that aim to better visualize the data as the framework has knowledge on the type of data.

For example, Castellanos-Garz\'{o}n and D\'{ı}az1 \cite{DNAVis} propose ``An Evolutionary and Visual Framework for Clustering of DNA Microarray Data'' which implements specific visualizations of the data that may not be useful when using other data than DNA data. Still, with this knowledge and this specific task, these visualizations may prove more useful than only the ones targeted at the general use-case of clustering. There are various examples for domains in which such specific frameworks exist, even for the task of clustering ``Memes'' \cite{Dang2017} to detect emerging events and topics driven by rumours, a framework has been created.

@@

In addition to visual frameworks for clustering, visual parameter space analysis is also relevant, as finding the best parameters for a clustering algorithm may be difficult. For such use-cases, Sedlmair et al. \cite{6876043} propose a framework that can help in finding parameters.

There are also a few tools that implement additional visualizations for consensus clustering. An example of such a tool is ConsensusClusterPlus \cite{10.1093/bioinformatics/btq170} which visualizes the co-association matrix created from the base clusterings. This matrix represents how often two points occur in a common cluster relative to the number of base clusterings. When looking at this as a heat map a clear block structure is supposed to indicate a good choice for the parameter $k$ which defines the number of clusters in the base results. It also implements Consensus Cumulative Distribution Function (CDF) Plots and Delta Area Plots, showing the CDF of the consensus matrix for each parameter $k$ and relative change for consecutive values of $k$. This Delta Area Plot is supposed to allow the user to determine a value $k$ at which the relative increase is negligible, similar to the elbow-method. Additional plots are also available, further enabling the user to compare the quality of consensus results regarding the parameter $k$.

@@

\section{Consensus Clustering}\label{sec:consensus}
Consensus clustering aims to combine multiple clustering solutions for the same data-set into one result. This result should be representative of the combined knowledge acquired by all base clusterings and therefore be more robust and accurate than the average result at least. As the consensus result should lie in the center of the solution space spanned by the base clusterings if there are sufficiently many good clusterings it should be able to overcome problems arising from outliers. In literature the problem of consensus clustering is often found under different names, those being cluster ensemble \cite{BOONGOEN20181}, clustering aggregation \cite{Gionis2005ClusteringA} and ensemble of partitions \cite{ensemblepartitions}. For the evaluation of the quality of a result, five main factors can be defined \cite{Ghaemi2009ASC,survey1}:\newline

@@

Consensus clustering includes two parts. Firstly base clusterings must be generated, which can be done with different strategies \cite{BOONGOEN20181}. One may choose to perform multiple runs of the same algorithm with the same parameters but different random seeds, the same algorithm with different parameters or even multiple different algorithms. All of these possibilities can be chosen while always clustering the full data, a different sub-sample of the data points (Bootstrapping) \cite{bootstrapping} each run (e.g. 80\%) or even different subspaces of the data. After the generation of base clusterings, the next step is to combine those with a consensus function. In general, consensus functions can be separated into two groups, Object Co-occurrence (O-Co) methods and Median Partitioning (MP) methods \cite{survey1}. There exist a large number of methods in both of those groups, which can be categorized according to what ideas they are based on.

\subsection{Object Co-occurrence}
Object Co-occurrence (O-Co) based consensus functions look at how points relate to each other and use this information to create a consensus result. If two points often occur in a common cluster of the base clusterings, intuition would suggest that they should also appear in a common cluster in the final result. The same holds for points most frequently clustered apart, indicating that they should not be clustered together by the consensus function. The different O-Co based methods formulate algorithms that aim to create a result reflecting these thoughts and usually run in a deterministic way, not seeing this task as an optimization problem. This contrasts the methodology of Median Partitioning methods as can be seen in the following section. To summarize the groups of O-Co consensus functions they are categorized by their underlying computational strategy. The different methods are based on:

@@

Relabeling and Voting based methods use the individual points as  ``voters'' whereby each point can vote for which cluster it should be assigned to \cite{4470298}. For this, to work, the different base clusterings need to map their clusters to one another, defining which cluster is associated most closely with which cluster from another base clustering. This mapping can be performed with algorithms like Kuhn’s Hungarian method \cite{Kuhn2010}.

@@

Graph and Hypergraph based methods create graphs from the base clusterings, representing the relationship between the points as edges. They then split the resulting graph into sub-graphs with graph partitioning \cite{Strehl:2003:CEK:944919.944935}, defining the clusters. Representative methods from this group are the Cluster-based Similarity Partitioning Algorithm (CSPA), the HyperGraph-Partitioning Algorithm (HGPA) and the Meta-CLustering Algorithm (MCLA). A variant of these are Minimum Spanning Tree algorithms which require a specific graph structure, a minimum spanning tee. This tree is created using Prim's Algorithm \cite[p.~276]{10.1007/978-3-540-85033-5_27} and the clusters are obtained by cutting edges. With $k$ edges cut the final result is derived through majority voting, a prominent method of this category is Divisive Clustering Ensemble with Automatic Cluster Number (DICLENS) \cite{6035671}. DICLENS is also able to estimate an optimal $k$ by computing all possible results and evaluating their quality.

Locally Adaptive Clustering Algorithms look at different subspaces of the data and compute weights from clustering results in these subspaces \cite{Domeniconi2007}. To create a final result, the computed clusterings and their weights are modeled as a graph, which is then partitioned.

@@

\subsection{Median Partitioning}
The main difference of Median Partitioning in comparison to Object Co-occurrence is that methods in this group tackle consensus clustering as an optimization problem. With the distance between clusterings defined the task of finding a median partition is to look for a solution that would be in the center of the solution space spanned by the base clusterings. In other words, the resulting clustering should have a minimal sum of distances to all base clusterings. As this problem has been shown to be NP-hard \cite{np_median_partition}, heuristics must be used. The different groups of median partitioning methods can be roughly described through the methods that they are based on:
%https://languagetool.org/

@@

Genetic Algorithm based methods require a distance measure and a fitness function to compute a consensus result \cite{Cristofor2002FindingMP}. Through evolutionary steps, different results are combined or modified until the fitness of the result is satisfactory or does not further improve over many iterations. At first, the results to be combined may be picked from the base clusterings or just be random assignments. For the choice of distance measure, Mirkin Distance is often used and comes with additional ideas on how genetic modifications should be applied \cite{5766165}, such that a good result is computed with higher probability. The Sum of Pairwise Distances can also be used with genetic algorithms. The assumption of this measure is that a good consensus result can be found by more strongly considering clusterings far apart in the solution space \cite{6694095}. The idea of using Kernel Functions was also proposed \cite{Vega-Pons:2010:WPC:1786814.1787121}, enabling faster computation regardless of the heuristic used, though mostly genetic algorithms are used with this as well.

@@

\part{Theory}
\chapter{Methods}\label{cha:methods}
The newly created tool uses a large number of existing methods that are described in this chapter. Here, I aim to give an overview of what exact methods were used, the theory behind them and in which scenarios they can be/are applied. For a more general overview of some of the topics relating the methods used, refer to the related work of this paper in Chapter \ref{cha:related_work}.

@@

\subsection{t-Distributed Stochastic Neighbor Embedding}
The dimensionality reduction method t-Distributed Stochastic Neighbor Embedding (t-SNE) \cite{Maaten2008VisualizingDU} works by converting the Euclidean distances between points into conditional probabilities that represent similarities. This similarity represents the probability that a point would be picked as the neighbor of another point if neighbors were picked proportionally to their probability density under a Student-t distribution centered at a given point. To layout the points, an initial randomly sampled solution is created and improved on iteratively with gradient descent, such that the similarities are best represented by the points in low-dimensional space. As inputs for the method, the number of dimensions $k$, the perplexity and some optimization parameters are needed. The number of resulting dimensions is defined by $k$ and the optimization parameters can be used to improve the runtime of the algorithm. The perplexity parameter can be seen as a smoothing parameter defining the effective number of neighbors for each point. This parameter is usually also fairly robust \cite{Maaten2008VisualizingDU}, good choices in practice are values between $5$ and $50$.

@@

\subsection{k-Means}
K-Means clustering is one of the most prominent methods used for clustering \cite{Jin2010}. The main implementations of k-Means are Lloyd's method \cite{4031353} and MacQueen's method \cite{macqueen1967} although important variants like k-Medians \cite{Dohan2015KmedianAT} and k-Medoids \cite{kmedoids}  are available.

The main idea of k-Means type clustering algorithms is to start with centroids which are points in the data-space representing the centers of clusters. Each point is therby assigned to the centroid closest to it according to Euclidean distance, defining the these clusters. To improve the result, the centroid is updated to the center of its corresponding points iteratively. This is repeated until no more changes happen or some convergence criterion is met. The update procedure of each iteration does not guarantee finding the optimum though, even if no more changes for the centroids can be observed. How the initial centroids are chosen may also impact the quality of the result greatly. To find a good initialization, Arthur and Vassilvitskii \cite{10.5555/1283383.1283494} proposed k-Means++ whereby the centroids are picked such that they are far apart with high probability. They showed that this kind of initialization makes the overall algorithm $O(log(k))$ competitive with the optimal clustering \cite{10.5555/1283383.1283494}. Still, the usual choice is to initialize them by picking random data points.

The main difference between Lloyd's and MacQueen's method is that Lloyd's method only recalculates the centroid positions every time an iteration is completed while MacQueen's method also updates them whenever a point changes its assigned clusters. This results in more computational cost during iterations, may allow for faster convergence though. The k-Medians method is different in regards to the distance metric used. Unlike Lloyd's and MacQueen's methods it does not use the Euclidean distance but the Manhattan-distance (1-Norm) to find the centroids. K-Medoids in contrast to all previous methods does not use centroids that may lie anywhere in the data space, but only existing points may be used. These are in this case called medoids and define the objects for which the average dissimilarity to all points within the cluster is minimal.

@@

\subsection{DBSCAN}
Density-Based Spatial Clustering of Application with Noise (DBSCAN) \cite{10.5555/3001460.3001507}, is an algorithm that unlike k-Means is not based on partitioning the data directly but on searching for dense regions in the data. For the calculation, two parameters are needed, namely epsilon ($\epsilon$) and a minimum number of objects ($\mu$ or MinPts). The parameter $\epsilon$ defines how large a region may be considered for neighborhood search and $\mu$ defines how many objects must be within that region for it to be considered dense. The following explanations are needed for a full understanding of the algorithm:
\newline

@@

Also, the relationship between $\epsilon$ and $\mu$ is interesting to observe in practice. Often, when trying different parameter sets for DBSCAN is suffices to fix one parameter to a somewhat sensible value while only sampling the other. If one for example knows that any region with less then $20$ ``close'' points should not be considered a cluster, it is possible to fix $\mu$ at $20$ and only sample $\epsilon$ to find a satisfactory result.

@@

The output of the OPTICS clustering algorithm is an ordered list of objects with their reachability-distances. To compute this list, all points are processed iteratively starting with any point initially. All points neighboring the current point are updated with their reachability distance from the current point, assigning the minimum of the new and previous values to the point. To select the next point the one with the smallest reachability value is chosen or a new random one if there are none. The resulting list of points with reachability distances can be visualized with a bar chart whereby valleys indicate clusters. As the definition of these valleys can be done in different ways, the result can be considered hierarchical.

>>>>

Clustering is the task of finding groups of points in data such that points within a group have some kind of relationship, while points in different groups do not relate. Defining how relationships can be expressed and how these groupings can be found are non-trivial tasks, for which reason a lot of research has been put towards this topic. Different ways of thinking about this problem, the expected results and the performance aspects of the algorithms resulted in a large number of different methods.

 To summarize which methods are available, many surveys were written over time \cite{7154919,1427769,7414675,surveyclustering}. Another aspect of clustering is which kinds of data can be clustered and even for more specific tasks like text clustering many methods are available \cite{5982288}. Even though these are relevant methods, they will not be mentioned further on, as this paper only works on processing numeric data. Also, work has been put into optimizing clustering algorithms for different types of architectures and machines, e.g. for distributed systems \cite{6322592} energy consumption and load balancing \cite{7586361}. Such aspects will also not be further considered, as this paper and the created tool do not aim to beat existing methods in runtime or energy efficiency.

To give an overview of existing clustering algorithms Xu and Tian \cite{surveyclustering} generalize the idea of algorithms to nine categories. These categories are defined by which fundamental ideas the algorithms are based on and stated in the following:
\begin{itemize}
  \item Partition
  \item Hierarchy
  \item Fuzzy Theory
  \item Distribution
  \item Density
  \item Graph Theory
  \item Grid
  \item Fractal Theory
  \item Model
\end{itemize}

Partition-based methods aim to find central points in the data to represent clusters. Data points are then assigned to the central point closest to them, whereby each central point forms a cluster. The most famous method in this group is k-Means which has been implemented in different ways; an example for a newer variant of it was proposed by Kanungo et al. \cite{1017616}.

Hierarchy-based methods at first consider each point as representing a cluster and merge groups of points to form clusters according to some logic. The opposite can also be done, first assuming all points belong to one cluster and then splitting them into multiple clusters. When starting with each point in one cluster, typical linking strategies are single link, joining clusters according to the distance between the closest points, average link, joining according to the average distance of all points from one cluster to the other, or maximum link, whereby the maximum distance between points of the clusters is considered.

Fuzzy Theory based methods do, unlike other methods, not produce discrete values for the cluster ownership label but a value in the interval $[0,1]$, representing how strongly a point belongs to a cluster. The usual method to compute such an ownership value is to optimize toward an objective function and seeing these values as the likelihood of a point belonging to a given cluster.

Distribution based methods aim to find clusters following the idea that any cluster should have its own mathematical distribution. In other words, each distribution found in the data should have a cluster with its generated points assigned to it. In most cases, Gaussian distributions are assumed, reducing the complexity of the search.

@@

Graph Theory based methods work by considering each point in the data as a vertex of a graph with their distances representing the edges. The algorithms then aim to partition the graph into $k$ sub-graphs, each representing a cluster, whereby the edges connecting the sub-graphs should have a large sum of the distances or small sum of similarities.

Grid based methods look at the data space as a multidimensional grid. The resulting rectangular spaces are then used in combination with hierarchical and density based methods to both improve the result and the computational performance of the algorithms. The cell-size of the grid allows a trade-off between those two goals.

Fractal Theory based methods aim to optimize the intrinsic quality of the fractal dimension \cite{surveyclustering}. In this context, fractal represents the geometry divisible into multiple parts having common characters with the whole \cite{surveyclustering}.

@@

%more on current development?
\section{Visual Frameworks}
Visual Frameworks are used to both allow the user analyzing data and empower him to produce better results. Often, when tackling a data analysis problem, the issue is not that there are no good methods available to solve it but the difficult choice of which method produces a satisfying result. As looking at quality values and data tables does often not reveal the necessary information for educated choices, visualizations and frameworks are needed. Note that for this paper the words framework, tool and application will be used synonymously as we consider them only differing in the way how the user uses the created work, whether it is to be used from code, through APIs or just as a program with a graphical user interface.

@@

For this reason, different frameworks have been proposed in literature, with a prominent one being Clustervision \cite{Kwon2018ClustervisionVS}. The Clustervision framework allows the user to visually explore the solutions created by multiple clustering algorithm runs, whereby each run may use a different algorithm and different settings for computation. Besides just showing the result and information on different properties of the resulting clusters, this framework ranks the computed results according to quality measures. As there exists a large number of different quality measures, each being biased towards/preferring specific cluster structures, the framework uses the average of five different quality measures as an indicator of the quality. A problem of this approach may occur when individual measures are biased and combined into a more general one. Then, one must be careful to not create new biases by the choice of how to combine them. The choices made by the authors of Clustervision \cite{Kwon2018ClustervisionVS}, for example, show favorable results for the presented use-cases but may be poor when facing a data-set that does not fit the assumptions of the chosen metrics. Aside from the problems arising from quality measures, Clustervision only considers clustering solutions that are at least $5\%$ different from other top results, possibly missing better solutions because they are similar to a good result. Also, by choosing a single result through this ranking process, the final choice can be in the best case as good as the best produced clustering, even though much more information is available when looking at all solutions. This left out knowledge can possibly be used to produce a combined result that may be better than any individual one resulting from simple clustering algorithms. These points are also the reason why the framework described later in this thesis was created.

VISTA \cite{VISTA} is another visual clustering framework which was proposed by Chen and Liu and consists of three parts. Firstly, VISTA allows for data filtering for visualization. In this step, data can be normalized, missing values can be handled and if the dimensionality is too high, reduction techniques can be used. Also, if the data is too large, it is possible to reduce it by extracting relevant sub-sets. Next, selecting and filtering of labels can be performed. After clustering, it is possible to partially extract cluster labels for the points using the framework's data filtering tools, allowing to only visualize some clusters, for example. Lastly, the tool allows for post-processing whereby a new cluster presentation is used which they name ClusterMap. ClusterMap projects the data on a two-dimensional grid showing the cluster labels in each cell and allowing for quick relabeling. Chen and Liu further iterate on this idea by introducing iVIBRATE \cite{10.1145/1148020.1148024} whereby they further analyze the accuracy of their tool and the theoretic background of their method.

Many clustering frameworks also implement visual elements such that clustering results or data can be explored visually, although this often only includes trivial plots. For example, the Java framework ELKI provides such simple functionality \cite{10.14778/2824032.2824115}. Generally, in most programming languages simple scatter plots and charts, which can be used whenever no further in-depth analysis is needed, are available.

\subsubsection{Data-set specific Frameworks}
On top of the general frameworks for visualizing clusterings, specific ones exist that aim to visualize the data better as the framework has knowledge on the type of data.

For example, Castellanos-Garz\'{o}n and D\'{ı}az1 \cite{DNAVis} propose ``An Evolutionary and Visual Framework for Clustering of DNA Microarray Data'' which implements specific visualizations of the data that may not be useful when using other data than DNA data. Still, with the knowledge about this specific task, these visualizations prove to be more useful than only the ones targeted at the general use-case of clustering. There are various examples for domains in which such specific frameworks exist; even for the task of clustering ``Memes'' \cite{Dang2017} to detect emerging events and topics driven by rumours, a framework has been created.

@@

In addition to visual frameworks for clustering, visual parameter space analysis is also relevant, as finding the best parameters for a clustering algorithm may be difficult. For such use-cases, Sedlmair et al. \cite{6876043} propose a framework that provides help in finding parameters.

There are also a few tools that implement additional visualizations for consensus clustering. An example is ConsensusClusterPlus \cite{10.1093/bioinformatics/btq170}, which visualizes the co-association matrix created from the base clusterings. This matrix represents how often two points occur in a common cluster relative to the number of base clusterings. When looking at this as a heat map, a clear block structure is supposed to indicate a good choice for the parameter $k$ which defines the number of clusters in the base results. It also implements Consensus Cumulative Distribution Function (CDF) Plots and Delta Area Plots, showing the CDF of the consensus matrix for each parameter $k$ and relative change for consecutive values of $k$. The Delta Area Plot is supposed to allow the user to determine a value $k$ at which the relative increase is negligible, similar to the elbow-method. Additional plots are also available, further enabling the user to compare the quality of consensus results regarding the parameter $k$.

@@

\section{Consensus Clustering}\label{sec:consensus}
Consensus clustering aims to combine multiple clustering solutions for the same data-set into one result. This result should be representative for the combined knowledge acquired by all base clusterings and therefore be more robust and accurate than the average result at least. As the consensus result should lie in the center of the solution space spanned by the base clusterings if there are sufficiently many good clusterings, it should be able to overcome problems arising from outliers. In literature, the problem of consensus clustering is often found under different names, those being cluster ensemble \cite{BOONGOEN20181}, clustering aggregation \cite{Gionis2005ClusteringA} and ensemble of partitions \cite{ensemblepartitions}. For the evaluation of the quality of a result, five main factors can be defined \cite{Ghaemi2009ASC,survey1}:\newline

@@

Consensus clustering includes two parts. Firstly, base clusterings must be generated, which can be done with different strategies \cite{BOONGOEN20181}. One may choose to perform multiple runs of the same algorithm with the same parameters but different random seeds, the same algorithm with different parameters or even multiple different algorithms. All of these possibilities can be chosen while always clustering the full data, a different sub-sample of the data points (Bootstrapping) \cite{bootstrapping} each run (e.g. 80\%) or even different subspaces of the data. After the generation of base clusterings, the next step is to combine those with a consensus function. In general, consensus functions can be separated into two groups, Object Co-occurrence (O-Co) methods and Median Partitioning (MP) methods \cite{survey1}. There exist a large number of methods in both of those groups, which can be categorized according to what ideas they are based on.

\subsection{Object Co-occurrence}
Object Co-occurrence (O-Co) based consensus functions look at how points relate to each other and use this information to create a consensus result. If two points often occur in a common cluster of the base clusterings, intuition would suggest that they should also appear in a common cluster in the final result. The same holds for points most frequently clustered apart, indicating that they should not be clustered together by the consensus function. The different O-Co based methods formulate algorithms that aim to create a result reflecting these thoughts and usually run in a deterministic way, not seeing this task as an optimization problem. This contrasts the methodology of Median Partitioning methods as can be seen in the following section. To summarize the groups of O-Co consensus functions, they are categorized by their underlying computational strategy. The different methods are based on:

@@

Relabeling and Voting based methods use the individual points as  ``voters'' whereby each point can vote for the cluster it should be assigned to \cite{4470298}. For this to work, the different base clusterings need to map their clusters to one another, defining which cluster is associated most closely with which cluster from another base clustering. This mapping can be performed with algorithms like Kuhn’s Hungarian method \cite{Kuhn2010}.

@@

Graph and Hypergraph based methods create graphs from the base clusterings, representing the relationship between the points as edges. They then split the resulting graph into sub-graphs with graph partitioning \cite{Strehl:2003:CEK:944919.944935}, defining the clusters. Representative methods from this group are the Cluster-based Similarity Partitioning Algorithm (CSPA), the HyperGraph-Partitioning Algorithm (HGPA) and the Meta-CLustering Algorithm (MCLA). A variant of these are Minimum Spanning Tree algorithms which require a specific graph structure, the minimum spanning tee. This tree is created using Prim's Algorithm \cite[p.~276]{10.1007/978-3-540-85033-5_27} and the clusters are obtained by cutting edges. With $k$ edges cut, the final result is derived through majority voting, a prominent method within this category is Divisive Clustering Ensemble with Automatic Cluster Number (DICLENS) \cite{6035671}. DICLENS is also able to estimate an optimal $k$ by computing all possible results and evaluating their quality.

Locally Adaptive Clustering Algorithms look at different subspaces of the data and compute weights from clustering results in these subspaces \cite{Domeniconi2007}. To create a final result, the computed clusterings and their weights are modeled as a graph, which is partitioned afterwards.

@@

\subsection{Median Partitioning}
The main difference of Median Partitioning in comparison to Object Co-occurrence is that methods in this group tackle consensus clustering as an optimization problem. With the distance between clusterings defined, the task of finding a median partition is to look for a solution that would be in the center of the solution space spanned by the base clusterings. In other words, the resulting clustering should have a minimal sum of distances to all base clusterings. As this problem has been shown to be NP-hard \cite{np_median_partition}, heuristics must be used. The different groups of median partitioning methods can be roughly described through the methods that they are based on:
%https://languagetool.org/

@@

Genetic Algorithm based methods require a distance measure and a fitness function to compute a consensus result \cite{Cristofor2002FindingMP}. Through evolutionary steps, different results are combined or modified until the fitness of the result is satisfactory or does not further improve over many iterations. At first, the results to be combined may be picked from the base clusterings or just be random assignments. For the choice of distance measure, Mirkin Distance is often used and comes with additional ideas on how genetic modifications should be applied \cite{5766165}, such that a good result is computed with higher probability. The Sum of Pairwise Distances can also be used with genetic algorithms. The assumption of this measure is that a good consensus result can be found by more strongly considering clusterings far apart in the solution space \cite{6694095}. The idea of using Kernel Functions was also proposed \cite{Vega-Pons:2010:WPC:1786814.1787121}, enabling faster computation regardless of the heuristic used, though mostly genetic algorithms are used in combination as well.

@@

\part{Theory}
\chapter{Methods}\label{cha:methods}
The newly created tool uses a large number of existing methods that are described in this chapter. Here, we aim to give an overview of what exact methods were used, the theory behind them and in which scenarios they can be/are applied. For a more general overview of some of the topics relating the methods used, refer to the related work of this paper in Chapter \ref{cha:related_work}.

@@

\subsection{t-Distributed Stochastic Neighbor Embedding}
The dimensionality reduction method t-Distributed Stochastic Neighbor Embedding (t-SNE) \cite{Maaten2008VisualizingDU} works by converting the Euclidean distances between points into conditional probabilities that represent similarities. This similarity represents the probability that a point would be picked as the neighbor of another point if neighbors were picked proportionally to their probability density under a Student-t distribution centered at a given point. To layout the points, an initial randomly sampled solution is created and improved on iteratively with gradient descent, such that the similarities are best represented by the points in low-dimensional space. As inputs for the method, the number of dimensions $k$, the perplexity and some optimization parameters are needed. The number of resulting dimensions is defined by $k$ and the optimization parameters can be used to improve the runtime of the algorithm. The perplexity parameter can be seen as a smoothing parameter defining the effective number of neighbors for each point. This parameter is usually fairly robust \cite{Maaten2008VisualizingDU}, good choices in practice are values between $5$ and $50$.

@@

\subsection{k-Means}
K-Means clustering is one of the most prominent methods used for clustering \cite{Jin2010}. The main implementations of k-Means are Lloyd's method \cite{4031353} and MacQueen's method \cite{macqueen1967}, although important variants like k-Medians \cite{Dohan2015KmedianAT} and k-Medoids \cite{kmedoids}  are available.

The main idea of k-Means type clustering algorithms is to start with centroids which are points in the data-space representing the centers of clusters. Each point is thereby assigned to the centroid closest to it according to Euclidean distance, defining the clusters. To improve the result, the centroid is updated to the center of its corresponding points iteratively. This is repeated until no more changes happen or some convergence criterion is met. The update procedure of each iteration does not guarantee finding the optimum though, even if no more changes for the centroids can be observed. How the initial centroids are chosen may also impact the quality of the result greatly. To find a good initialization, Arthur and Vassilvitskii \cite{10.5555/1283383.1283494} proposed k-Means++ whereby the centroids are picked such that they are far apart with high probability. They showed that this kind of initialization makes the overall algorithm $O(log(k))$ competitive with the optimal clustering \cite{10.5555/1283383.1283494}. Still, the usual choice is to initialize them by picking random data points.

The main difference between Lloyd's and MacQueen's method is that Lloyd's method only recalculates the centroid positions every time an iteration is completed while MacQueen's method also updates them whenever a point changes its assigned clusters. This results in more computational cost during iterations, may allow for faster convergence though. The k-Medians method is different in regards to the distance metric used. Unlike Lloyd's and MacQueen's methods it does not use the Euclidean distance but the Manhattan-distance (1-Norm) to find the centroids. K-Medoids in contrast to all previous methods does not use centroids that may lie anywhere in the data space, but only existing points may be used. These are called medoids and define the objects for which the average dissimilarity to all points within the cluster is minimal.

@@

\subsection{DBSCAN}
Density-Based Spatial Clustering of Application with Noise (DBSCAN) \cite{10.5555/3001460.3001507} is an algorithm, that unlike k-Means, is not based on partitioning the data directly but on searching for dense regions in the data. For the calculation, two parameters are needed, namely epsilon ($\epsilon$) and a minimum number of objects ($\mu$ or MinPts). The parameter $\epsilon$ defines how large a region may be considered for neighborhood search and $\mu$ defines how many objects must be within that region for it to be considered dense. The following explanations are needed for a full understanding of the algorithm:
\newline

@@

Also, the relationship between $\epsilon$ and $\mu$ is interesting to observe in practice. Often, when trying different parameter sets for DBSCAN it is sufficient to fix one parameter to a somewhat sensible value while only sampling the other. If one, for example, knows that any region with less then $20$ ``close'' points should not be considered a cluster, it is possible to fix $\mu$ at $20$ and only sample $\epsilon$ to find a satisfactory result.

@@

The output of the OPTICS clustering algorithm is an ordered list of objects with their reachability-distances. To compute this list, all points are processed iteratively, starting with any point initially. All points neighboring the current point are updated with their reachability distance from the current point, assigning the minimum of the new and previous values to the point. To select the next point, the one with the smallest reachability value or a new random one if there are none reachable is chosen. The resulting list of points with reachability distances can be visualized with a bar chart whereby valleys indicate clusters. As the definition of these valleys can be done in different ways, the result can be considered hierarchical.

\section{Consensus Functions}
Consensus functions have the goal of merging multiple different clustering solutions together such that the overall result represents a consensus between all base clusterings. In this work, we focused on two methods that aim to achieve this.
\subsection{Co-association Matrix based Consensus}
Co-association Matrix based consensus functions \cite{Monti2003} first create a co-association matrix that indicates how often two points $p$ and $q$ appear in a common cluster, i.e. if $p$ and $q$ are clustered together $70\%$ of the time the corresponding entry of the matrix is $0.7$. To extract a consensus result from this matrix different methods can be applied. One possibility is to choose a clustering algorithm that takes this matrix as an input, converts it to a dissimilarity/distance measure and performs clustering with it. Another possibility is to use linking of sets strategies, which is also done for the implementation in the tool. This has the benefit of being trivial to implement and not imposing a cluster structure, as it would, for example, be the case when using k-Means. The possible linking strategies are single-link, using the similarity of the two points of the sets where it is maximal, average-link, using the average similarity between all point pairs, or maximum link, whereby the smallest similarity is used. With each of these strategies, the pair of sets for which this linking value is maximal (i.e. they are most similar) is merged. To stop the merging process and obtain the final result, either a threshold value can be set for the link value, indicating from where on no more merging should be performed or a number of target clusters $k$ can be defined, stopping when only $k$ sets remain. Another option for defining when to stop merging is the Lifetime Criterion. For the implementation of the tool, the average-link strategy is chosen per default with either defining $k$ or using this Lifetime Criterion to estimate $k$.
